Evaluating infectious disease forecasts with allocation scoring rules

### 2023-12-05 alloscore_manuscript/alloscore-application.Rnw
updates to manuscript with start at multi-k analysis.

In the example of this one week, JHUAPL-Bucky had a worse allocation score than CU-select because its forecasts led to allocations that sent excess resources to several states, such as Ohio, Pennsylvania, and Michigan, that would have been more effectively allocated to states that did receive enough resources to meet their needs, such as Florida and California (Figure \ref{multi-loc-ranks}A). These allocation errors resulted because that model’s forecasts did not consistently capture the relative resource needs across different states. (Figure \ref{fig:thermometer-plot} \& \ref{multi-loc-ranks}A). The CU-select model made some similar errors — most prominently, over-allocating resources to Ohio — but overall, it did a better job of forecasting the relative resource demands across different locations.

On the other hand, CU-select had worse performance as measured by WIS. Its forecasts were biased downwards, and it consistently incurred a large penalty for underprediction (Figure \ref{fig:multi-loc-rank}B). Predictions from the JHUAPL-Bucky model were wider, and included the observed level of daily hospital admissions more often. Therefore, WIS did not as often assign that model severe penalties for forecasts that underpredicted or overpredicted the actual hospitalization count.

@@

Using the integrated allocation score summarizes allocation scores across a range of possible values of the constraint ($K$) (Section \ref{sec:methods.detailed.integrated_allocation}).
This could be useful in situations where the actual constraint may not be known, or could only be specified as a probability distribution.

 - something here from Ben's analysis?

<<ias-comparison, results='asis'>>=
## for visual context
## curve(dnorm(x, mean=15000, sd=3000), from=0, to=30000)
ks <- seq(5000, 25000, by=200)
kprobs <- data.frame(K = ks,
                     weight = dnorm(ks, mean=15000, sd=3000)/sum(dnorm(ks, mean=15000, sd=3000)))

@@

p1 <- ggplot(left_join(ias, as), aes(x=AS, y=IAS)) + geom_point(color="blue")
p2 <- ggplot(left_join(ias, as), aes(x=AS, y=IAS_wt)) + geom_point(color="red")

p3 <- slim_alloscore_2021.12.20 |>
  group_by(model, K) |>
  summarize(`Allocation score` = sum(components)) |>
  left_join(kprobs) |>
  ggplot(aes(x=K)) +
  geom_point(aes(y=weight), color = "red") +
  geom_point(y=1/length(unique(slim_alloscore_2021.12.20$K)), color="blue") +
  geom_vline(xintercept = 15000, color="black")

cowplot::plot_grid(p2, p3, p1,
                   nrow=2,
                   align="v",
                   labels = "AUTO")

left_join(ias, as) |>
  xtable::xtable() |>
  print(include.rownames = FALSE)
@

<<multi-k, fig.height=4>>=
slim_alloscore_2021.12.20 |>
  group_by(model, K) |>
  summarize(`Allocation score` = sum(components)) |>
  mutate(model = factor(model, levels = ias$model)) |>
  ggplot(aes(x=K, y=`Allocation score`, color=model, linetype=model)) +
  geom_line()
@

>>>>

In the example of this one week, JHUAPL-Bucky had a worse allocation score than CU-select because its forecasts led to allocations that sent excess resources to several states, such as Ohio, Pennsylvania, and Michigan, that would have been more effectively allocated to states that did receive enough resources to meet their needs, such as Florida and California (Figure \ref{fig:multi-loc-ranks}A). These allocation errors resulted because that model’s forecasts did not consistently capture the relative resource needs across different states. (Figure \ref{fig:thermometer-plot} \& \ref{fig:multi-loc-ranks}A). The CU-select model made some similar errors — most prominently, over-allocating resources to Ohio — but overall, it did a better job of forecasting the relative resource demands across different locations.

On the other hand, CU-select had worse performance as measured by WIS. Its forecasts were biased downwards, and it consistently incurred a large penalty for underprediction (Figure \ref{fig:multi-loc-ranks}B). Predictions from the JHUAPL-Bucky model were wider, and included the observed level of daily hospital admissions more often. Therefore, WIS did not as often assign that model severe penalties for forecasts that underpredicted or overpredicted the actual hospitalization count.

@@

The integrated allocation score summarizes allocation scores across a range of possible values of the constraint ($K$), possibly taking weights into account for different values of $K$ that might be more or less likely (Section \ref{sec:methods.detailed.integrated_allocation}).
This could be useful in situations where the actual constraint may not be known, or could only be specified as a probability distribution.

This also underscores the points above that allocation scores are highest where the observed value is hear the allocation constraint, but here we are holding the date/models fixed and varying K.

Allocation scores are fairly stable across a range of values for K, and therefore IAS is not very sensitive to the choice of distribution on K, and does not differ much from the regular old allocation score as long as that allocation score was computed at a reasonable choice of K.

Interpretation of Figure \ref{fig:multi-k} needed.

<<ias-comparison>>=
## for visual context
## curve(dnorm(x, mean=15000, sd=3000), from=0, to=30000)
ks <- seq(5000, 25000, by=200)
kprobs <- data.frame(K = ks,
                     weight = dnorm(ks, mean=15000, sd=3000)/sum(dnorm(ks, mean=15000, sd=3000)))

@@

ias_to_plot <- left_join(ias, as) |>
  pivot_longer(cols = c(IAS, IAS_wt), names_to = "IAS_type", values_to = "IAS") |>
  mutate(IAS_type = ifelse(IAS_type=="IAS", "uniform", "centered at 15,000"))

p1 <- ggplot(ias_to_plot, aes(x=AS, y=IAS, color=IAS_type, shape=IAS_type)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE, linetype=2) +
  scale_color_manual(name = "Integrated Allocation Score", values = c("#386cb0", "#fdc086")) +
  scale_shape_manual(name = "Integrated Allocation Score", values = c(1, 2)) +
  geom_abline(slope=1, intercept=0, color="gray") +
  theme(legend.position = c(0.05,.9), legend.justification = c(0,1),
        legend.background=element_rect(fill = alpha("white", 0.2)))

p2 <- slim_alloscore_2021.12.20 |>
  group_by(model, K) |>
  summarize(`Allocation score` = sum(components)) |>
  left_join(kprobs) |>
  ggplot(aes(x=K)) +
  geom_point(aes(y=weight), color = "#386cb0") +
  geom_point(y=1/length(unique(slim_alloscore_2021.12.20$K)), color="#fdc086", shape = 2) +
  geom_vline(xintercept = 15000, color="gray")
@

<<multi-k-scores, results='asis'>>=
left_join(ias, as) |>
  rename(`IAS centered at 15k` = IAS_wt,
         `IAS uniform` = IAS) |>
  xtable::xtable(caption = "Allocation scores shown in Figure \\ref{fig:multi-k}C, sorted by Integrated Allocation Scores (IAS) centered at 15,000.", digits = 0) |>
  print(include.rownames = FALSE, )
@

<<multi-k, fig.height=10, fig.width = 8, fig.cap = "Allocation scores across different resource constraints (K) for the 10 models that made forecasts on December 20, 2021. Panel A shows, for each model, the allocation score assigned for values of K between 200 and 60,000 at increments of 200. The allocation scores show a sharp peak just under 20,000 which was the eventually observed number of hospitalizations. It is hardest to get allocations exactly right, with no excess unmet need, when the resource constraint is exactly equal to the number of resources needed. Panel B shows two possible weighting functions for the Integrated Allocation Score (IAS). The first (dark blue circles) computes weights proportional to a normal distribution centered at 15,000 (solid vertical gray line) with a standard deviation of 3,000, and truncated to be between 5,000 and 25,000. The second (light orange triangles) uses a uniform weight for all possible values of K. Note that the Allocation Score (AS), used in earlier sections of the application, only uses the allocation scores at the `pre-defined' value of K=15,000. the  Panel C shows how either method of computing the IAS (y-axis, again in blue circles and orange squares) is correlated with AS (x-axis, computed at K=15,000). Every point represents the AS and IAS for one model. Note that the IAS centered at 15,000 are much more closely correlated  with the AS values, which makes sense as this version of the IAS assigns higher weights to values of the IAS that are closer to the single K at which the AS scores were computed.">>=
p3 <- slim_alloscore_2021.12.20 |>
  group_by(model, K) |>
  summarize(`Allocation score` = sum(components)) |>
  mutate(model = factor(model, levels = rev(ias$model))) |>
  ggplot(aes(x=K, y=`Allocation score`, color=model, linetype=model)) +
  geom_line() +
  theme(legend.position = c(0.01,.99), legend.justification = c(0,1),
        legend.background=element_rect(fill = alpha("white", 0.2)),
        legend.title = element_blank(), legend.text = element_text(size=7))

cowplot::plot_grid(p3, p2, p1,
                   nrow=3,
                   align = "h",
                   labels = "AUTO")
@
